\subsection{XML provider for external representation}

XML providers are used to store and retrieve parameter configurations of
operators and workflows.
As a consequence, (typically) transient data are not stored (and retrieved).

\icode{ALDXMLObjectType} is the base type of all Alida XML data types.
In only contains the classname (fully qualified classname).
Extending types will add the data itself.
Such XML representations do not only contain the data itself,
but allow to infer the class of the represented object. 
Thus, it is feasible to instantiate the correct object and fill it with all
data as stored in the XML object.

Currently there exist the following extending types:

\begin{tabular}{l|l}
\icode{ALDXMLArrayType} & to hold 1D arrays and collections \\
\icode{ALDXMLParametrizedType} & for parametrized classes holding a list of
(key,value)-pairs,\\
& where the values are of type \icode{ALDXMLObjectType} \\
\icode{ALDXMLEnumType} & for an enum value, which is just represented by a string \\
\icode{ALDXMLAnyType} & to hold any type, used, e.g., for primitive data
types,\\ & numerical data types
\end{tabular}

For writing an XML provider on your own
\begin{enumerate}
\item	create an XML beans object which can hold the data item.
	This may be a native XML beans data type like \icode{XmlInt} or
	\icode{XmlString}, or a custom XML beans data type typically created via an
	XML schema and the XML bean compiler (\icode{xmlbean} ant task).
\item	create an object of type \icode{ALDXMLAnyType},
	\begin{itemize}
	\item	set its member \icode{'className'} to the class name of the data item
	\item	set its value to the XML beams object created in step 1.
	\end{itemize}
\end{enumerate}
